\paperTitle

### 2014-05-18 doc/icml2014/undirected.tex
Updated undirected section

While we now have a complete algorithm for estimating directed and undirected
models, optimizing the full likelihood (\equationref{logLinearSupervised})
can still be computationally intractable for undirected models with high treewidth
due to the intractability of the log-partition function $A(\theta)$.
One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
but these generally lead to inconsistent estimates of $\theta$.
We thus turn to an older idea of pseudolikelihood \citep{besag75pseudo}.
The pseudolikelihood objective is a sum over the log-probability of each
variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
  L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p_{\theta^*}}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood
provides consistent estimates which are computationally efficient
but less statistically efficiency.
(Coincidentally, this is the same high-level motivation for using method of
moments in the first place.)

@@

If we knew the marginals for each neighborhood,
\begin{align}
  \mu_{a,\sN(a)} &\eqdef \E[\phi_{a,\sN(a)}(a,\sN(a))], \label{eqn:neighborhoodFeatures}
\end{align}
then we would be able to optimize the pseudolikelihood objective again without
having access to any labeled data.
However, $\{a\}\cup\sN(a)$ does not always have exclusive views
(for example, consider $a = h_1$ and $\sN(a) = \{h_2,h_3,h_4\}$ in \figureref{examples-tree}).

However, we can decompose $\{a\}\cup\sN(a)$ as follows:
conditioning on $a$ partitions $\sN(a)$ into independent subsets;
let $\sB(a)$ be the collection of these subsets,
which we will call \emph{sub-neighborhoods}.
For example, $\sB(h_1) = \{ \{h_2\},\{h_3\},\{h_4\} \}$ in \figureref{examples-tree}
and $\sB(h_{2,2}) = \{ \{ h_{1,2}, h_{2,3}, h_{3,2}, h_{2,1} \} \}$ contains a single sub-neighborhood.

A key observation is that for each $B \in \sB(a)$, each $\{a\} \cup B$ is bidependent:
conditioning on $a$ does not introduce new independencies within $B$ by construction of $\sB(a)$,
and conditioning on any $b \in B$ does not either since every other $b' \in B \backslash \{b\}$
is connected to $a$.
Assuming $\sG$ is bottlenecked, by \lemmaref{bottleneck-views},
we have that $\{a\} \cup B$ has exclusive views.
Hence, we can recover $\Pr(a,B)$ for each $a$ and $B \in \sB(a)$.
Based on conditional independence of the sub-neighborhoods $B$ given $a$,
we have that $\Pr(a, \sN(a)) = \Pr(a) \prod_{B \in \sB(a)} \Pr(B \mid a)$.
This allows us to estimate the expected features $\mu_{a,\sN(a)}$ and use these quantities
in the optimization of the pseudolikelihood objective.

@@

Note that our pseudolikelihood-based approach
does depend exponentially on the size of the sub-neighborhoods,
which could be larger than the largest clique size.
Therefore,
for this approach to be effective,
each node essentially should have low degree or locally exhibits a lot of
conditional independence.
%While this method is only viable for graphs with small sub-neighboof low
%degree, i.e.\@ when $|\sN(a)|$ is small,
On the positive side, we can handle graphical models with high treewidth;
neither sample nor computational complexity depends on the treewidth.
For example, in the grid model, the treewidth is linear in the size of the grid,
but it is the degree (only 4 in this case) that governs the computational complexity.

>>>>

While we now have a complete algorithm for estimating directed and undirected
models, optimizing the full likelihood (\equationref{logLinearSupervised})
can still be computationally intractable for undirected models with high treewidth
due to the intractability of the log-partition function $A(\theta)$.
One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
but these generally lead to inconsistent estimates of $\theta$.
We thus turn to an older idea of pseudolikelihood \citep{besag75pseudo}.
The pseudolikelihood objective is a sum over the log-probability of each
variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
  L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p_{\theta^*}}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood
provides consistent estimates which are computationally efficient
but less statistically efficiency.\footnote{
Coincidentally, this is the same high-level motivation for using method of
moments in the first place.}

@@

If we knew the marginals for each neighborhood,
\begin{align}
  \mu_{a,\sN(a)} &\eqdef \E[\phi_{a,\sN(a)}(a,\sN(a))], \label{eqn:neighborhoodFeatures}
\end{align}
then we would be able to optimize the pseudolikelihood objective again without
having access to any labeled data.
Unfortunately, $\{a\}\cup\sN(a)$ does not always have exclusive views.
For example, consider $a = h_1$ and $\sN(a) = \{h_2,h_3,h_4\}$ in \figureref{examples-tree}.

However, we can decompose $\{a\}\cup\sN(a)$ as follows:
conditioning on $a$ partitions $\sN(a)$ into independent subsets;
let $\sB(a)$ be the collection of these subsets,
which we will call \emph{sub-neighborhoods}.
For example, $\sB(h_1) = \{ \{h_2\},\{h_3\},\{h_4\} \}$ in \figureref{examples-tree}
and $\sB(h_{2,2}) = \{ \{ h_{1,2}, h_{2,3}, h_{3,2}, h_{2,1} \} \}$ contains a single sub-neighborhood in \figureref{examples-mrf}.

A key observation is that for each sub-neighborhood $B \in \sB(a)$, each $\{a\} \cup B$ is bidependent:
conditioning on $a$ does not introduce new independencies within $B$ by construction of $\sB(a)$,
and conditioning on any $b \in B$ does not either since every other $b' \in B \backslash \{b\}$
is connected to $a$.
Assuming $\sG$ is bottlenecked, by \lemmaref{bottleneck-views}
we have that $\{a\} \cup B$ has exclusive views.
Hence, we can recover $\Pr(a,B)$ for each $a$ and $B \in \sB(a)$.
Based on conditional independence of the sub-neighborhoods $B$ given $a$,
we have that $\Pr(a, \sN(a)) = \Pr(a) \prod_{B \in \sB(a)} \Pr(B \mid a)$.
This allows us to estimate the expected features $\mu_{a,\sN(a)}$ and use these quantities
in the optimization of the pseudolikelihood objective.

@@

Note that our pseudolikelihood-based approach
does depend exponentially on the size of the sub-neighborhoods,
which could be larger than the largest clique size.
Therefore,
for this approach to be effective,
each node essentially should have low degree or locally exhibit a lot of
conditional independence.
%While this method is only viable for graphs with small sub-neighboof low
%degree, i.e.\@ when $|\sN(a)|$ is small,
On the positive side, we can handle graphical models with high treewidth;
neither sample nor computational complexity depends on the treewidth.
For example, in the grid model, the treewidth is linear in the size of the grid,
but it is the degree (only 4 in this case) that governs the computational complexity.
